### 2019-08-27 Manuscript/Manuscript.tex

Indonesia and New Guinea have an outsized influence on modern chemical weathering fluxes. The confluence of steep topography, a warm and wet tropical climate, and the presence of mafic lithologies results in high fluxes of Ca and Mg cations and associated \COtwo consumption \citep{Gaillardet1999a, Hartmann2009a, Hartmann2014a}. There has been a significant increase of subaerially-exposed land area within the region since the Miocene associated with ongoing arc-continent collision between Australia and the Sunda-Banda arc system (\citealp{Molnar2015a, Hall2017a}; Fig. \ref{fig:shoreline_growth}). The collision between the Australian-Indian and Eurasian plates and intervening arc terranes has led to the emergence of Indonesia and New Guinea and the exhumation of large supra-subduction ophiolites and associated forearcs in the warm, wet tropics \citet{Macdonald2019a}.

@@

Along with the Neogene increase in subaerially-exposed land area and tectonic topography associated with arc-continent collision, there has also been a decrease in subaerially-exposed land area on the Sunda Shelf \citep{Sarr2019a}. Large portions of this low-relief area of continental crust were likely emergent throughout the Neogene, and were repeatedly subaerially exposed during the Pleistocene as recently as the Last Glacial Maximum \citep{Hall2002a}. Similarities in terrestrial biotas of Borneo and mainland Southeast Asia suggest the existence of land bridges on the Sunda Shelf throughout the Miocene \citep{Moss1998a}. During the Eocene through early Miocene, the rifting of the South China Sea resulted in subsidence of the Sunda Shelf \citep{Morley2013a} and recent subsidence is interpreted to be a result of dynamic topography, triggered by subduction-related mantle flow \citep{Sarr2019a}. In the Miocene, this region exposed a mixture of felsic basement rocks and siliciclastic sediments \citet{Darmadi2007a}. The relatively low-relief and felsic lithologies of the region contrasts with the high-relief and mafic lithologies of most of Indonesia and New Guinea.

@@

Of the 3,339 unique parameter combinations, the ``removed Indonesia and New Guinea'' scenario resulted in steady-state \pCOtwo's of 598-729~ppm (5-95 percentile values), the ``reduced Indonesia and New Guinea'' scenario of 547-638~ppm, and the ``reduced Indonesia and New Guinea + Sunda Shelf'' of 533-611~ppm (Fig. \ref{fig:scenario_pCO2}). These results suggest that without the Neogene increase in subaerially-exposed land area and tectonic topography associated with arc-continent collision, pre-Industrial \pCOtwo would have been \textgreater500~ppm. Notably, the effect of the Sunda Shelf on steady-state \pCOtwo is minimal, given the low Mg+Ca of the lithologies that comprise the Sunda Shelf as well as its low slope.

%%%
% maybe place this at the end of the discussion
The movement of the Deccan Traps into the tropical rain belt together with the low-latitude emplacement of the Ethiopian Traps is an alternative hypothesis that has been put forth to explain the draw down of \COtwo leading into Oligocene glaciation of Antarctica \citep{Kent2008a, Kent2013a}. To explore the sensitivity of global weatherability and climate state to the presence of these traps, we test three additional scenarios using GEOCLIM (Fig. \ref{fig:scenario_pCO2}). In the ``removed Deccan Traps'' and the ``removed Ethiopian Traps'' scenarios, all volcanics associated with the respective traps are replaced with bulk continental crust. In the ``removed both traps'' scenario, all volcanics associated with both traps are replaced with bulk continental crust. These scenarios result in steady-state \pCOtwo's \textless500~ppm, distinctly lower than those for the three Indonesia and New Guinea scenarios.
%%%

@@

Indonesia and New Guinea are potent sinks of carbon because they contain large swaths of ophiolites that are exposed along E-W belts of tectonic topography in the warm wet tropics. This paleogeographical feature originated in the Middle Miocene with the collision between the Outer Melanesian arc and the distal margin of Australia (59, 61, 62). In New Guinea, this resulted in the exhumation of the 1440 km long Irian-Marum-April ophiolite belt (Fig. 3; suture 7). The Middle Miocene (16–14 Ma) basal Markats Formation contains siliciclastic sediment that was transported from the south into the forearc basin associated with the New Guinea ophiolite belt (61, 63). This precedes, by several million years, the beginning of widespread synorogenic sedimentation to the south and on the Australian continental basement at ca. 12 Ma (59). Mountain building began in the Late Miocene, ca. 8-7 Ma (59, 62) with 3-4 km of denudation between 8-5 Ma (64). Exhumation of the Central Range accelerated over the past 4 Myrs due to slab-breakoff and buoyant uplift under New Guinea (61,65). After arc-continent collision in New Guinea, a left-lateral strike-slip margin developed on the northern margin (62). Over the past 4 Myr, this change in plate motion caused the westward extrusion and exhumation of the Cretaceous to Oligocene ophiolites of Obi (67), Halmahera (68, 69), and northern New Guinea (70) (Fig. 3; sutures 8-10). In eastern New Guinea, progressive jamming of the north-dipping subduction zone has caused major uplift over the past 4 Ma (van Ufford & Cloos, 2005), which is well dated with a change in provenance from continental to volcanic detritus (Abbott et al., 1994; Abbott, 1995) and unroofing of basement-cored uplifts (Hill & Gleadow, 1989).

>>>>

Indonesia and New Guinea have an outsized influence on modern chemical weathering fluxes. The confluence of steep topography, a warm and wet tropical climate, and the presence of mafic lithologies results in high fluxes of Ca and Mg cations and associated \COtwo consumption \citep{Gaillardet1999a, Hartmann2009a, Hartmann2014a}. There has been a significant increase of subaerially-exposed land area within the region since the Miocene associated with ongoing arc-continent collision between Australia and the Sunda-Banda arc system (\citealp{Molnar2015a, Hall2017a}; Fig. \ref{fig:shoreline_growth}). The collision between the Australian-Indian and Eurasian plates and intervening arc terranes has led to the emergence of Indonesia and New Guinea and the exhumation of large supra-subduction ophiolites and associated forearcs in the warm, wet tropics \citep{Macdonald2019a}.

@@

Along with the Neogene increase in subaerially-exposed land area and tectonic topography associated with arc-continent collision, there has also been a decrease in subaerially-exposed land area on the Sunda Shelf \citep{Sarr2019a}. Large portions of this low-relief area of continental crust were likely emergent throughout the Neogene, and were repeatedly subaerially exposed during the Pleistocene as recently as the Last Glacial Maximum \citep{Hall2002a}, potentially driving climate amplification on orbital timescales (ADD DENEZIO&TIERNY REFS). Similarities in terrestrial biotas of Borneo and mainland Southeast Asia suggest the existence of land bridges on the Sunda Shelf throughout the Miocene \citep{Moss1998a}. During the Eocene through early Miocene, the rifting of the South China Sea resulted in subsidence of the Sunda Shelf \citep{Morley2013a} and recent subsidence is interpreted to be a result of dynamic topography, triggered by subduction-related mantle flow \citep{Sarr2019a}. In the Miocene, this region exposed a mixture of felsic basement rocks and siliciclastic sediments \citet{Darmadi2007a}. The relatively low-relief and felsic lithologies of the region contrasts with the high-relief and mafic lithologies of most of Indonesia and New Guinea.

@@

Using the 3,339 unique parameter combinations, the ``removed Indonesia and New Guinea'' scenario resulted in steady-state \pCOtwo's of 598-729~ppm (5-95 percentile values), the ``reduced Indonesia and New Guinea'' scenario of 547-638~ppm, and the ``reduced Indonesia and New Guinea + Sunda Shelf'' of 533-611~ppm (Fig. \ref{fig:scenario_pCO2}). These results suggest that without the Neogene increase in subaerially-exposed land area and tectonic topography associated with arc-continent collision, pre-Industrial \pCOtwo would have been $sim$500-700~ppm. Notably, the effect of the Sunda Shelf on steady-state \pCOtwo is relatively small, given the low Mg+Ca of the lithologies that comprise the Sunda Shelf as well as its low slope.

These GEOCLIM \pCOtwo values resemble proxy-based estimates for the early Miocene \citep{Foster2017a}, indicating that the increase in subaerially-exposed land area and tectonic topography of Indonesia and New Guinea is enough to explain climate change over the Neogene. Furthermore, our paleoshoreline reconstructions (Fig. \ref{fig:shoreline_growth}) show a broad increase in exposed land area over the Neogene and an acceleration over the past 4~m.y. This trend mirrors broad cooling over the Neogene and accelerated cooling over the past 4~m.y. that culminated in the initiation of the Northern Hemisphere ice sheets.

%%%
% maybe place this at the end of the discussion
The movement of the Deccan Traps into the tropical rain belt together with the low-latitude emplacement of the Ethiopian Traps is an hypothesis that has been put forth to explain the draw down of \COtwo leading into Oligocene glaciation of Antarctica \citep{Kent2008a, Kent2013a}. To explore the sensitivity of global weatherability and climate state to the presence of these traps, we test three additional scenarios using GEOCLIM (Fig. \ref{fig:scenario_pCO2}). In the ``removed Deccan Traps'' and the ``removed Ethiopian Traps'' scenarios, all volcanics associated with the respective traps are replaced with bulk continental crust. In the ``removed both traps'' scenario, all volcanics associated with both traps are replaced with bulk continental crust. These scenarios result in steady-state \pCOtwo's \textless500~ppm, distinctly lower than those for the three Indonesia and New Guinea scenarios.
%%%

@@

Indonesia and New Guinea are potent sinks of carbon because they contain large swaths of ophiolites that are exposed along E-W belts of tectonic topography in the warm wet tropics. In New Guinea, this paleogeographical feature originated in the Middle Miocene with the collision between the Outer Melanesian arc and the distal margin of Australia (59, 61, 62), which resulted in the exhumation of the 1440 km long Irian-Marum-April ophiolite belt. The Middle Miocene (16–14 Ma) basal Markats Formation contains siliciclastic sediment that was transported from the south into the forearc basin associated with the New Guinea ophiolite belt (61, 63). This precedes, by several million years, the beginning of widespread synorogenic sedimentation to the south and on the Australian continental basement at ca. 12 Ma (59). Mountain building began in the Late Miocene, ca. 8-7 Ma (59, 62) with 3-4 km of denudation between 8-5 Ma (64). Exhumation of the Central Range accelerated over the past 4 Myrs due to slab-breakoff and buoyant uplift under New Guinea (61,65). After arc-continent collision in New Guinea, a left-lateral strike-slip margin developed on the northern margin (62). Over the past 4 Myr, this change in plate motion caused the westward extrusion and exhumation of the Cretaceous to Oligocene ophiolites of Obi (67), Halmahera (68, 69), and northern New Guinea (70) (Fig. 3; sutures 8-10). In eastern New Guinea, progressive jamming of the north-dipping subduction zone has caused major uplift over the past 4 Ma (van Ufford & Cloos, 2005), which is well dated with a change in provenance from continental to volcanic detritus (Abbott et al., 1994; Abbott, 1995) and unroofing of basement-cored uplifts (Hill & Gleadow, 1989).
